---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

Load lattice library and set locale to English.
```{r results = "hide"}
library(lattice)
Sys.setlocale("LC_TIME", "C")
```

## Loading and preprocessing the data
Load the data.
```{r}
# unzip file if it doesn't exist yet
if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}

# read in csv file
activity <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
Calculate the total number of steps taken per day ignoring missing values.
```{r}
# take sum of steps factorized by date
steps <- tapply(activity$steps, activity$date, function(x) sum(x, na.rm = TRUE))
```

Make a histogram of the total number of steps taken each day.
```{r}
hist(steps, xlab = "Number of steps", main = "Histogram")

steps_mean <- round(mean(steps), digits = 2)
steps_median <- median(steps)
```
Mean of total number of steps taken per day: `r steps_mean`  
Median of total number of steps taken per day: `r steps_median`


## What is the average daily activity pattern?
Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
```{r}
# take mean of steps factorized by interval
interval_steps <- tapply(activity$steps, activity$interval, function(x) mean(x, na.rm = TRUE))

# plot time series of average steps per interval with custom x-axis ticks
plot(interval_steps, type = "l", xlab = "Interval", xaxt="n", ylab = "Number of steps")
custom_axTicks <- axTicks(1)
custom_axTicks[1] <- 1
axis(1, at = axTicks(1), labels = names(interval_steps[custom_axTicks]))

# calculate interval with most steps
max_interval <- names(which(interval_steps == max(interval_steps)))
```
5-minute interval containing maximum number of steps: `r max_interval`

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r}
missing_sum <- sum(is.na(activity))
```
Total number of missing values in the dataset: `r missing_sum`  
  
The strategy for filling in all of the missing values in the dataset is to take the mean of a 5-minute interval and assign it to a missing value in that interval.

```{r}
# copy dataset
activity_full <- activity

# loop over all rows and fill in missing values of steps with average of the interval
for (i in 1 : nrow(activity_full)) {
    if (is.na(activity_full$steps[i])) {
        activity_full$steps[i] <- interval_steps[as.character(activity_full$interval[i])]
    }
}
```

Make a histogram of the total number of steps taken each day.
```{r}
steps_full <- tapply(activity_full$steps, activity_full$date, function(x) sum(x, na.rm = TRUE))
hist(steps_full, xlab = "Number of steps", main = "Histogram")

steps_full_mean <- round(mean(steps_full), digits = 2)
steps_full_median <- round(median(steps_full), digits = 2)
```
Mean of total number of steps taken per day: `r as.character(steps_full_mean)`  
Median of total number of steps taken per day: `r as.character(steps_full_median)`

The mean and median from imputed step counts differ from the mean and median where missing values were removed because less steps got taken into account. Imputing missing values has the impact of increasing the total daily number of steps.


## Are there differences in activity patterns between weekdays and weekends?
For this part the dataset with the filled-in missing values is used.  
  
Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
daytype <- vector("integer")

# loop over all rows and assign daytype either 0 for weekdays or 1 for weekends
for (i in 1 : nrow(activity_full)) {
    if (is.element(weekdays(as.Date(activity_full$date[i])), c("Saturday", "Sunday"))) {
        daytype[i] <- 0
    } else {
        daytype[i] <- 1
    }
}

# create factor column daytype in data frame
activity_full$daytype <- factor(daytype, labels = c("weekend", "weekday"))
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
# create two datasets for weekend and weekday data
activity_weekend <- subset(activity_full, daytype == "weekend")
activity_weekday <- subset(activity_full, daytype == "weekday")

# take mean of steps factorized by interval
interval_steps_weekend <- tapply(activity_weekend$steps, activity_weekend$interval, mean)
interval_steps_weekday <- tapply(activity_weekday$steps, activity_weekday$interval, mean)

# get maximum average step count of both weekends and weekdays
max_average_value <- max(c(interval_steps_weekend, interval_steps_weekday))

# draw xyplot factorized by daytype, plotting average steps over interval
xyplot(steps ~ interval | daytype, data = activity_full, xlab = "Interval", ylab = "Number of steps", layout = c(1, 2), type = "l", ylim = c(0, max_average_value + max_average_value/10), index.cond = list(c(2, 1)), panel = function(x, y, ...) { panel.average(x, y, horizontal = FALSE, col = "blue", ...)})
```

There is a difference in activity patterns between weekdays and weekends. The number of steps on weekdays is greater in early and late intervals while the number of steps on the weekends is increased during the day.